Case-based Arabic Morphological Analyzer for Arabic Text Mining

### 2010-12-20 tex/rp.tex
lit review and rebuttle 1

{\bf Goals.} 
CATARM tends to relational Arabic text mining using 
{\bf (1.)} a novel case-based morphological analyzer.
{\bf (2.)} CATARM will build a novel and efficient 
linguistic computational model (LCM) of Arabic. 
%The LCM will be implemented using efficient structures. 
The LCM will be amenable to {\bf (3.)} Arabic specific simplifications 
inferred from the correspondence between the formal Arabic rules
and the relational user queries, and 
{\bf (4.)} to simplifications 
inferred from statistical features extracted from the documents 
using language independent techniques such as 
hidden Markov models (HMM)  and 
latent semantic indexing~\cite{LSI89}.
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (5.)} CATARM extracts semantic graphs 
and uses set arithmetic and graph analysis techniques
to establish relations between the graphs. {\bf (5.)} CATARM will 
solve the case studies 
presented in Sections~\ref{s:design:lit}--\ref{s:design:med}.

@@

CATARM will discover relations between Arabic text documents that 
are hard to discover by humans and will have direct live and
cost saving benefits in cases similar to the above mentioned 
case studies. 
CATARM will qualitatively enhance research in the cultural and 
literature fields. 
It may also help investors analyze Arabic news and data from Arab 
markets prior to making important decisions.

@@

The directed acyclic graph (DAG) 
in Figure~\ref{f:narrators} shows a partial order relation (POR) between
narrators.
We automatically extracted the POR from hadith text 
documents.
The nodes in boxes are the \RL{matn} of the hadith, 
and the other nodes are the narrators.
Given a set of chains of narrators similar to 
Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
that merged equal names in the chains. 
This partial order graph is instrumental to automate
localizing narrators in biography documents and
segmenting biography documents.

@@

The Lead PI worked in the area of software Arabization since 1996 
as he worked on developing Arabization modules for Windows NT 
and an Arabic string manipulation library~\cite{Zar96}. 
The Lead PI is also involved in working on an open source 
project that aims at comprehensive automation of authenticity 
checks against a corpus of literature heritage~\cite{Zar06}.
The Lead PI is leading an effort at AUB to introduce 
Computational Arabic in the curriculum of the ECE program. 
He is also leading a team of students in an effort to place 
the infrastructure needed for this project to start.

@@

IE targets forming data structures and instances of these 
data structures out of a collection of documents.
IE tries to fit the output of IR and NLP to templates of 
interest to the user.
Data mining can be applied at the end on the output of the 
IE process to answer user queries about the input 
documents~\cite{JHa05}.

@@

\subsection{Literature authentication case study}
\label{s:design:lit}

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.9\columnwidth}{!}
{\includegraphics{figs/literature_case-study1.pdf}
} 
\caption{CATARM diagram for the literature case study }
\label{f:literature}
}
\end{figure}

The diagram in Figure~\ref{f:literature} describes our 
high level vision for what the literature case study
will look like in the CATARM framework. 
The boxes represent 
computational blocks provided by CATARM, 
ovals represent data blocks which are mainly the 
inputs and the outputs of the computational blocks. 
Dotted data blocks represent intermediate results,
and solid and shaded blocks represent the initial input 
and the final output respectively.

@@

@@

We think that we can discover such features by building
a linguistic computational model of the Arabic language
based on its formal grammar and formal morphology
as defined in several Arabic linguistic 
sources~\cite{Sha73,Abd00,Abd001}.
We will build our morphological analyzer on top 
of this linguistic model in addition to the current existing
concatenative-based analysis techniques.

Let $L_q$ be the query language grammar, 
$L_A$ be the grammar that expresses the Arabic linguistic
computational model, and let $q$ be a user query.
We compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We will explore methods that use $\rho_q$ to simplify 
$L_A$ before we start the morphological analysis. 
We think the possible simplifications that we can perform
here are Language specific as they will reflect the 
correspondence between the formal query language definition $L_q$, 
the usage patterns of $L_q$  by Arabic users, 
and $L_A$ the formal LCM for Arabic.

@@

{\bf Case-based morphological analyzer.}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSM with linear execution time using 
the trie~\cite{Aoe:89} data structure.

@@

{\bf Linguistic computational model of Arabic.}
We think that a linguistic computational model for the Arabic
language will perform better in applications where
relations between Arabic text documents are involved. 
Such a model helps capture features unique and specific 
to the Arabic language and make use of that in the analysis.

@@

@@

{\bf Statistical components.}
We will make use and build statistical components that can be used
in CATARM. 
We will need statistical components to rank solutions, compute 
probable decisions, estimating good thresholds.

{\bf Graph algorithms.}
The relations extracted from the text documents will often 
be expressed in terms of graphs.
We will need textbook graph algorithms such as traversals, 
topological sorts, spanning trees, strongly connected 
components, minimum flow  and other algorithms.
For example, in the literature case study, the discovery
of articulation vertices may point to important narrators 
who may have an insignificant number of narrations 
but still connect narrator communities.

{\bf Query solver.}
This is a logical solver that will solve query expressions to \
resolve matching text and relations.
We will need to use existing off the shelf constraint solvers 
and set theory satisfiability solvers to address our queries. 
The solver should return and rank approximate matches.
That is if a three entities out of four required in a relation are
found, the solution is still worth reporting.

The solver will call the morphological analyzer to compute 
morphological matches to the atomic elements of the query. 
Then each operation in the query such as a set intersection
operation will correspond to a computational component. 
The solver will call these components and integrate their
results.

The solver may make use of existing 
propositional~\cite{MiniSAT04} or
higher order logic~\cite{Z308} solvers if needed. 
These solvers can solve satisfiability questions.
We will explore the feasibility and utility 
of extracting a necessary set of 
constraints from the query and answering that with an off-the-shelf
logic solver.

{\bf Case studies.}
We will work on developing the necessary components to complete 
the solution for the hadith literature  case study. 
We will also work on the security, health and network traffic case studies.
While working on the case studies, we will add the components
we build to our library of components. 
This will enrich CATARM with a useful set of computational 
components.

{\bf Visualization.} 
We will build a visualization tool that allows fast 
and guided inspection and correction mechanism. 
The CATARM visualization aid will be a tag and relation color
sensitive browser.
The interface will allow a fast and guided inspection and 
correction mechanism.

@@

We detailed an aggressive work plan in Table~\ref{t:workplan}. 
The plan spans over three years and projects a timeline
of work per quarter. 
The row headers describe work items, and the column headers
describe quarters in the corresponding year of the project.
The legends {\bf P, W, D, M, } 
and {\bf E} 
describe the status of the items 
and stand for 
{\bf P:} {\em planning phase}, 
{\bf W:} {\em work in progress}, 
{\bf D:} {\em deliverable}, 
{\bf M:}  {\em maintenance}, 
and 
{\bf E: } {\em enhance} respectively.

The $1,2,3,$ and $4$
superscripts of the deliverables in the 
in case studies row refer to the four 
case studies described in the proposal.

We will use open source tools, libraries, and frameworks to 
build our libraries.
We will build on existing open source programming and text 
mining tools as well as open source Arabic text analysis tools 
based on NLP such as MADA+TOKAN~\cite{Rot08}, 
NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.

@@

We will evaluate CATARM against the case studies
that we discussed in Section~\ref{s:design} and 
that are listed below. 
\begin{enumerate}
\item Hadith literature
\item Security
\item Health
\item Network content
\end{enumerate}

@@

\subsection{CATARM evaluation }

@@

The team will publish the results of the research in the form 
of refereed conference and journal papers as well as tool papers 
addressing each of the tools. 
The team will also hold workshops to educate possible users on the 
tools once the tools are available.
The Arabic text analysis framework will be partially based on 
existing work in the open source domain and will thus be available 
as an open source tool. 
Interested researchers and users will be able to download the 
tools for free and use them for non-profit non-commercial 
purposes.

>>>>

{\bf Goals.} 
CATARM tends to relational Arabic text mining using 
{\bf (1.)} a novel case-based morphological analyzer.
{\bf (2.)} CATARM will build a novel and efficient 
linguistic computational model (LCM) for Arabic. 
%The LCM will be implemented using efficient structures. 
The LCM will be amenable to novel 
{\bf (3.)} Arabic specific simplifications 
inferred from the correspondence between the formal Arabic rules
and the user queries, and 
{\bf (4.)} language independent simplifications
inferred from statistical features %extracted from the documents 
using language independent techniques such as 
hidden Markov models (HMM)~\cite{JanHMM06}
and latent semantic indexing (LSI)~\cite{LSI89}.
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (5.)} CATARM extracts semantic graphs 
and uses set arithmetic and graph analysis techniques
to establish relations between the graphs. 
{\bf (6.)} CATARM will solve the case studies 
presented in Sections~\ref{s:design:lit}--\ref{s:design:med}.

@@

CATARM will discover relations between Arabic text documents that 
are hard to discover by humans and will have direct live and
cost saving benefits.% in cases similar to the above mentioned 
%case studies. 
CATARM will qualitatively enhance research in the cultural and 
literature fields. 
It may also help investors analyze Arabic news and data from Arab 
markets prior to making important decisions.
CATARM may be used to flag spammers 
posting biased reviews in Arabic web forums.

@@

The directed acyclic graph (DAG) 
in Figure~\ref{f:narrators} shows a partial order relation (POR) 
between narrators.
We automatically extracted the POR from three books of hadith 
selected arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}
\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
The nodes in boxes are the \RL{matn} of the hadith, 
and the other nodes are the narrators.
Given a set of chains of narrators similar to 
Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
that merged equal names in the chains. 
This partial order graph is instrumental to automate
localizing narrators in biography documents and
segmenting biography documents.

@@

The Lead PI worked in the area of software Arabization since 1996 
as he worked on developing Arabization modules for Windows NT 
and an Arabic string manipulation library~\cite{Zar96}. 
The Lead PI is also involved in working on an open source 
project that aims at comprehensive automation of authenticity 
checks against a corpus of hadith literature~\cite{Zar06}.
Volunteers including experts in the field of Islamic hadith
are involved in the project.
The Lead PI is leading an effort at AUB to introduce 
Computational Arabic in the curriculum of the ECE program. 
He is also leading a team of students in an effort to place 
the infrastructure needed for this project to start.

@@

IE targets forming data structures and instances of these 
data structures out of a collection of documents.
IE tries to fit the output of IR and NLP to templates of 
interest to the user.
%Data mining can be applied at the end on the output of the 
%IE process to answer user queries about the input 
%documents~\cite{JHa05}.

@@

We describe CATARM using four case studies and then move
to describe the different tools needed to build
CATARM.

\subsection{Hadith authentication case study}
\label{s:design:lit}

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.9\columnwidth}{!}
{\includegraphics{figs/literature_case-study1.pdf}
} 
\caption{CATARM diagram for the hadith case study }
\label{f:literature}
}
\end{figure}

The diagram in Figure~\ref{f:literature} describes our 
high level vision for what the hadith case study
will look like in the CATARM framework. 
The boxes represent 
computational blocks provided by CATARM, 
ovals represent data blocks which are mainly the 
inputs and the outputs of the computational blocks. 
Dotted data blocks represent intermediate results,
and solid and shaded blocks represent the initial input 
and the final output respectively.

@@

Toutanova in~\cite{Tou09} suggested that using 
a joint model to do lemmatization and POS tagging performs better
than the direct and the pipelined approach. 
We take the first argument further to suggest a full joint NLP task 
model via using the NLP query to simplify and render the analysis
more accurate at the low level morphological analysis.

Later, Toutanova, in joint work with Smith and others~\cite{Smi10},
suggested better extraction quality from comparable or related
corpora. 
They consider two sets of documents of the same nature from
two different languages and extract parallel structures. 
We are actually considering the same genre of analysis in 
our proposal and case studies only in the same language. 
For example, the hadith narrations are comparable to the hadith 
biographies as far as the researcher is concerned with authenticity.
The same applies to the clinical reports and the pharmaceutical 
records, and the investigative reports from the crime scenes
and the interrogative reports with suspects.

@@

We think that we can discover such features by building
a linguistic computational model (LCM) of the Arabic language
based on its formal grammar and formal morphology
as defined in several Arabic linguistic 
sources~\cite{Sha73,Abd00,Abd001}.
We will build our morphological analyzer on top 
of this LCM addition to the current existing
concatenative-based analysis techniques.

Our LCM will differ from ElixirFM~\cite{Otakar:07}
in that it will follow a relational approach rather 
than a functional morphological approach. 
That is the solutions will be expressed in a directed acyclic graph 
(DAG)
representing a hierarchy of relations between the morphemes 
rather than moprhological trees mapping the morphemes to the 
solutions.
We think this is better since it provides a more succint
representation for the solutions and because interesting relations 
between morphemes can be computed statically before
the input is presented to the analyzer.
Morphological trees are traversals of these DAGs, are expensive 
to exhaustively compute and store statically, 
and are expensive to compute dynamically. 
Our intuition is that Arabic allows relational strcutures
such as noun sentences and complex hierarchical entities.

Let $L_q$ be the query language grammar, 
$L_A$ be the grammar that expresses the Arabic linguistic
computational model, and let $q$ be a user query.
We compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We will explore methods that use $\rho_q$ to simplify 
$L_A$ before we start the analysis. 
We think the possible simplifications that we can perform
here are Language specific as they will reflect the 
correspondence between the formal query language definition $L_q$, 
the usage patterns of $L_q$  by Arabic users, 
and $L_A$ the formal LCM for Arabic.

@@

{\bf Case-based morphological analyzer.~~}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSM with linear execution time using 
the trie~\cite{Aoe:89} data structure.

@@

{\bf Linguistic computational model of Arabic.~~}
We think that a linguistic computational model for the Arabic
language will perform better in applications where
relations between Arabic text documents are involved. 
Such a model helps capture features unique and specific 
to the Arabic language and make use of that in the analysis.

@@

In ~\cite{Dao09}, Daoud observes that the pipeline model,
where morphology is performed first and syntactic and other
processes follow, is not adequate for Arabic. 
He suggests a synchronized syntactic and morphological
approach.  
We take the argument further to 
allow a tighter integration between the several pipeline
levels up to the NLP query in question by providing a 
case-based morphological analyzer, a linguistic computational
model of Arabic, and a query language that can all be expressed
as FSMs. 
The work~\cite{AlSham08} reports a similar result where
the automated addition of syntax knowledge improved steeming
in terms of accuracy and efficiency.

Here we point to a subtletie that shows how complex 
low level Arabic morphology is. 
The work in~\cite{Dao09} assumes that a noun
cannot be followed by a preposition in Arabic and uses that 
to resolve an ambiguity in the 
example~\noTrNoVocRL{.s-a.hb-aa s-amy _dhb-aa il_A alswq} 
(Sami's two companions went to the market) 
    and decides that \noTrNoVocRL{_dhb-aa} 
is a verb (they went)
and not a noun since it is followed by \noTrNoVocRL{il_A}.
One can 
say~\noTrNoVocRL{a_h_d  .s-a.hb-aa s-amy _dhb-aa il_A alswq} 
(Sami's two companions took gold to the market) and 
now \noTrNoVocRL{_dhb-aa} is a noun (gold stressed with double 
diactric).
This shows that even complex and advanced techniques such as
~\cite{Dao09} can be misled by ambiguous compatibility
rules.

@@

{\bf Statistical components.~~}
We will make use and build statistical components that can be used
in CATARM. 
We will need statistical components to rank solutions, compute 
probable decisions, estimating good thresholds.

{\bf Graph algorithms.~~}
The relations extracted from the text documents will often 
be expressed in terms of graphs.
We will need textbook graph algorithms such as traversals, 
topological sorts, spanning trees, strongly connected 
components, minimum flow  and other algorithms.
For example, in the hadith literature case study, the discovery
of articulation vertices may point to important narrators 
who may have an insignificant number of narrations 
but still connect narrator communities.

{\bf Query solver.~~}
The query solver considers query expressions and resolves them 
into matching text and relations. 
The solver returns and ranks approximate matches.
That is if a three entities out of four required in a relation are
found, the solution is still worth reporting.

The solver passes the query to a front end that builds
a parse graph of the query. 
It uses the parse graph to simplify the LCM and the MA
as discussed before. 
Then it traverses the resulting parse graph. 
When the solver meets an atomic element of the query, 
it will call the morphological analyzer to compute 
morphological matches to the atomic element.
When the solver meets an operation of the query, 
it will call the corresponding computational component 
and passes to it the solutions computed so far from children
nodes. 
The solver will use existing off the shelf constraint solvers 
and set theory satisfiability solvers to address the queries. 
The solver may make use of existing 
propositional~\cite{MiniSAT04} or
higher order logic~\cite{Z308} solvers if needed. 
These solvers can solve satisfiability questions.

We explain the ranking aspect of the solver with a simple
example. 
Consider $q$ the parse graph of a query with $n$ nodes where 
each node corresponds to either a detected entity or
to an operation that can be computed with one of CATARM components. 
We define the rank of a solution to be the ratio of  the 
satisfied nodes in $q$ over the possible satisfied
nodes in $q$.

%We will explore the feasibility and utility 
%of extracting a necessary set of 
%constraints from the query and answering that with an off-the-shelf
%logic solver.

{\bf Case studies.~~}
We will work on developing the necessary components to complete 
the solution for the hadith literature  case study. 
We will also work on the security, health and network traffic case studies.
While working on the case studies, we will add the components
we build to our library of components. 
This will enrich CATARM with a useful set of computational 
components.

{\bf Visualization.~~}
We will build a visualization tool that allows fast 
and guided inspection and correction mechanism. 
The CATARM visualization aid will be a tag and relation color
sensitive browser.
The interface will allow a fast and guided inspection and 
correction mechanism.

@@

We detailed an aggressive work plan in Table~\ref{t:workplan}. 
The plan spans over three years each with four quarters. 
The row headers describe work items, and the column headers
describe quarters in the corresponding year of the project.
The legends {\bf P, W, D, M, } 
and {\bf E} 
describe the type of activity taking place 
and stand for 
{\bf P:} {\em plan}, 
{\bf W:} {\em work }, 
{\bf D:} {\em deliver}, 
{\bf M:}  {\em maintain}, 
and 
{\bf E: } {\em enhance} respectively.

The $1,2,3,$ and $4$
superscripts of the deliverables in the 
row of case studies refer to the four 
case studies described in the proposal.

Work can happen on all items in parallel. 
Planing and research work can start on items that depend
on other items. 
For example, the LCM optimization items will be under 
active planing and do not need to be totally idle 
while the LCM is being built. 
That will allow better chances to build the LCM data structures
in a flexible way that requires less modifications if any
when the optimizations take place.

That said, we describe the dependency of the items on 
each other as follows. 
The work on language optimizations and data optimizations 
depend on the Arabic LCM item. 
The work on language optimizations depends on the design
of the query language as well. 
The work on the query solver depends on the design of the 
query language. 
The entity and relation extractors depend on the morphological 
analyzer. 
However, they can start since we have already
developed an working analyzer as a preliminary step. 
The rest of the items can all work in parallel.

We will use open source tools, libraries, and frameworks to 
build our componenets.
We will build on existing open source programming and text 
mining tools as well as open source Arabic text analysis tools 
based on NLP such as MADA+TOKAN~\cite{Rot08}, 
NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.

@@

We will evaluate CATARM against the case studies
that we discussed in Section~\ref{s:designmethods}
and that are listed below. 
\begin{enumerate}
\item Hadith literature
\item Security
\item Health
\item Network content
\end{enumerate}

@@

\subsection{External evaluation }

@@

The team will publish the results of the research in the form 
of refereed conference and journal papers as well as tool papers 
addressing each of the tools. 
The team will also hold workshops to educate possible users on the 
tools once the tools are available.
Graduate students such as Masters and PhD students are examplary t
target audience for such workshops.
The Arabic text analysis framework will be partially based on 
existing work in the open source domain and will thus be available 
as an open source tool. 
Interested researchers and users will be able to download the 
tools for free and use them for non-profit non-commercial 
purposes.
